Some examples in extended logic counting classes

### 2017-01-09 LICS/beyond.tex

As a example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extended with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L$ is in $\qfo(\tc)$, where $\chi_L(x) = 1$ if $x \in L$, and $\chi_L(x) = 0$ otherwise. Thus, if we assume that $\qfo(\tc)$ captures $\shl$ (over ordered structures), then we have that $\chi_L \in \nlog$ for every $L \in \nlog$. This implies that $\nlog = \ulog$,\footnote{A decision language $L$ is in $\ulog$ is there exists a logarithmic-space NTM $M$ accepting $L$ and satisfying that $\tma_M(x) = 1$ for every $x \in L$.} and thus contradicts the widely believed assumption that $\ulog \subsetneq \nlog$.

>>>>

As a example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extended with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L$ is in $\qfo(\tc)$, where $\chi_L(x) = 1$ if $x \in L$, and $\chi_L(x) = 0$ otherwise. Thus, if we assume that $\qfo(\tc)$ captures $\shl$ (over ordered structures), then we have that $\chi_L \in \nlog$ for every $L \in \nlog$. This implies that $\nlog = \ulog$,
\footnote{A decision language $L$ is in $\ulog$ is there exists a logarithmic-space NTM $M$ accepting $L$ and satisfying that $\tma_M(x) = 1$ for every $x \in L$.} 
and thus contradicts the widely believed assumption that $\ulog \subsetneq \nlog$.

### 2017-01-09 LICS/conclusions.tex

Here comes the conclusions.

>>>>

We proposed a framework based on Weighted Logics to develop a descriptive complexity theory for complexity classes of functions.
%In particular, we show how this framework can be used to capture fundamental counting complexity classes such as $\fp$, $\shp$ and $\fpspace$, among others. Moreover, we use it to define a hierarchy inside $\shp$, identifying counting complexity classes with good closure and approximation properties, and which admit natural complete problems. Finally, by adding recursion to the framework, we show how to capture lower counting complexity classes such as $\shl$.	
We consider the results of this paper as a first step in this direction.
%towards a descriptive complexity theory for complexity classes of functions. ç
In this sense, there are several directions for future research, some of which are mentioned here. 
$\totp$ is an important counting complexity class for which we are missing a logical characterization. In the same way, we are missing logical characterization of harder fundamental complexity class such as $\spanl$. We would also like to define a larger syntactic subclass of $\shp$ where each function admits an FPRAS; notice that $\cpm$ is an important problem admitting an FPRAS\cite{JSV04} that is not included in the classes defined in Section \ref{sec-clo}. Moreover, by following the approach proposed in
\cite{I83}, we would like to include second-order free variables in the operator for counting paths introduced in Section \ref{sec:beyond}, so to have alternative ways to capture $\fpspace$ and even $\shp$. Finally, the least fixed point operator introduced in Section \ref{sec:beyond} clearly deserves further investigation.

### 2017-01-09 LICS/counting.tex

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} is straightforward to prove, and it follows by the fact that $\shp$ is closed under first- and second-order sum.
Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
\begin{corollary} \label{prop:capture-shP-LFP}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}
By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
%The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
%So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
%By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
%Indeed, 
%By using free second-order variables one can easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and 
%By only using free first-order variables one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
%(e.g. $2^{n}$), so
Hence, it is not  clear how to capture
%impossible 
%to capture 
$\fp$ 
%or $\fpspace$ 
by following the approach proposed in~\cite{SalujaST95}. 
%if we follow previous approaches like
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.
% the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
%built in the ouptape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.
%That is, without these operators is difficult to capture $\fp$.

>>>>

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} is straightforward to prove, and it follows by the fact that $\shp$ is closed under first- and second-order sum.
%Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
%\begin{corollary} \label{prop:capture-shP-LFP}
%	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
%\end{corollary}

By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
%The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
%So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
%By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
%Indeed, 
%By using free second-order variables one can easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and 
%By only using free first-order variables one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
%(e.g. $2^{n}$), so
Hence, it is not  clear how to capture
%impossible 
%to capture 
$\fp$ 
%or $\fpspace$ 
by following the approach proposed in~\cite{SalujaST95}. 
%if we follow previous approaches like
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.
% the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
%built in the ouptape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.
%That is, without these operators is difficult to capture $\fp$.

### 2017-01-09 LICS/extentions.tex

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{fenner1994gap} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. Then this can be easily done in $\qsoz$ with the following formula:
	\[
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	\]
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
Interestingly, this is exactly what we need to capture  $\gp$.
\begin{proposition} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{proposition}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

>>>>

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{fenner1994gap} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. Then this can be easily done in $\qsoz$ with the following formula:
	\[
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	\]
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{corollary} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{corollary}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

### 2017-01-09 LICS/saluja.tex

\begin{figure*}
\begin{center}
\begin{tikzpicture}
\node[rectw] (n1) {$\E{0}$};
\node[rectw, right=0.5cm of n1] (n2) {};
\node[rectw, above=0.5cm of n2] (n3) {$\E{1}$}
	edge[draw=white] node {\rotatebox{45}{$\subsetneq$}} (n1);
\node[rectw, below=0.5cm of n2] (n4) {$\QE{0}$}
        edge[draw=white] node {\rotatebox{315}{$\subsetneq$}} (n1);
\node[rectw, right=0.5cm of n2] (n5) {$\QE{1}$}
        edge[draw=white] node {\rotatebox{315}{$\subsetneq$}} (n3)
        edge[draw=white] node {\rotatebox{45}{$\subsetneq$}} (n4);
\node[rectw, right=0.5cm of n5] (n6) {$\U{1}$}       
        edge[draw=white] node {$\subsetneq$} (n5);
\node[rectw, right=0.5cm of n6] (n7) {$\QU{1}$}       
        edge[draw=white] node {$=$} (n6);
\node[rectw, right=0.5cm of n7] (n8) {$\E{2}$}       
        edge[draw=white] node {$\subsetneq$} (n7);
\node[rectw, right=0.5cm of n8] (n9) {$\QE{2}$}       
        edge[draw=white] node {$=$} (n8);        
\node[rectw, right=0.5cm of n9] (n10) {$\U{2}$}       
        edge[draw=white] node {$\subsetneq$} (n9);
\node[rectw, right=0.5cm of n10] (n11) {$\QU{2}$}       
        edge[draw=white] node {$=$} (n10); 
\node[rectw, right=0.5cm of n11] (n12) {$\sfo$}       
        edge[draw=white] node {$=$} (n11); 
\end{tikzpicture}
\end{center}

\caption{The relationship between the $\sfo$-hierarchy and the $\eqso(\fo)$-hierarchy, where $\E{1}$ and $\QE{0}$ are incomparable. \label{fig-sfo-eqso}}
\end{figure*}

>>>>

\begin{figure*}
\begin{center}
\begin{tikzpicture}
\node[rectw] (n1) {$\E{0}$};
\node[rectw, right=0.5cm of n1] (n2) {};
\node[rectw, above=0.4cm of n2] (n3) {$\E{1}$}
	edge[draw=white] node {\rotatebox{45}{$\subsetneq$}} (n1);
\node[rectw, below=0.4cm of n2] (n4) {$\QE{0}$}
        edge[draw=white] node {\rotatebox{315}{$\subsetneq$}} (n1);
\node[rectw, right=0.5cm of n2] (n5) {$\QE{1}$}
        edge[draw=white] node {\rotatebox{315}{$\subsetneq$}} (n3)
        edge[draw=white] node {\rotatebox{45}{$\subsetneq$}} (n4);
\node[rectw, right=0.5cm of n5] (n6) {$\U{1}$}       
        edge[draw=white] node {$\subsetneq$} (n5);
\node[rectw, right=0.5cm of n6] (n7) {$\QU{1}$}       
        edge[draw=white] node {$=$} (n6);
\node[rectw, right=0.5cm of n7] (n8) {$\E{2}$}       
        edge[draw=white] node {$\subsetneq$} (n7);
\node[rectw, right=0.5cm of n8] (n9) {$\QE{2}$}       
        edge[draw=white] node {$=$} (n8);        
\node[rectw, right=0.5cm of n9] (n10) {$\U{2}$}       
        edge[draw=white] node {$\subsetneq$} (n9);
\node[rectw, right=0.5cm of n10] (n11) {$\QU{2}$}       
        edge[draw=white] node {$=$} (n10); 
\node[rectw, right=0.5cm of n11] (n12) {$\sfo$}       
        edge[draw=white] node {$=$} (n11); 
\end{tikzpicture}
\end{center}
\vspace{-0.4cm}
\caption{The relationship between the $\sfo$-hierarchy and the $\eqso(\fo)$-hierarchy, where $\E{1}$ and $\QE{0}$ are incomparable. \label{fig-sfo-eqso}}
\end{figure*}
